### 2020-05-06 probabilisticProgrammingMartingales.tex
Reorganised proof of Theorem 1 and related developments.

\newcommand{\Real}{\mathbb{R}}
\newcommand\expect[1]{\mathbb{E}[#1]}
\newcommand\set[1]{\{#1\}}

@@

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}

@@

\begin{theorem} \label{rankable implies ast}
\changed[lo]{If a closed SPCF term is rankable by $f$, then $(f(M_n))_{n \in \omega}$ is a supermartingale adapted to the filtration $(\mathcal{F}_n)_{n \in \omega}$ where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.}
\end{theorem}

%\begin{proof}
Given a closed term $M$ and a ranking function $f$ for it, define random variables on the probability space $S$ (where $s$ is a random variable) by
\begin{align*}
(M_n,s_n) & = \red^n(M,s) \\
y_0 & = 0 \\
y_{n+1} & = \min \{ k \mid k>y_n, M_k \text{ a value or of the form } E[\tY N] \}\\
M'_n & = M_{y_n} \\
X_n & = f(M'_n)
\end{align*}
and define a filtration $\mathcal{F}_n = \sigma(M_k, k \leq n)$ (i.e. all the samples used up to step $n$).

@@

\lo{Take $M \in \Lambda^0$ (closed SPCF terms).
Define, for each $n \in \omega$, the random variable $M_n : S \to \Lambda^0$ by $M_n(s) := \pi_0 (\red^n(M, s))$.

Given $M \in \Lambda^0$, say that $M$ is \emph{type-1} (respectively 2, 3 and 4) if $M$ has the shape $E[\tY \lambda x. N]$ (respectively $E[\tsample]$, $E[R]$ where $R$ is any other redex, and of a value).
By abuse of language, we refer to the three respective types of redexes as type-$1$, 2 and 3.

Henceforth fix $M \in \Lambda^0$.
For $i \in \{1, 2, 3\}$ and $n \in \omega$, define function $f_{i, n+1}(M) : S \to \Real$ by
\[
f_{i, n+1}(M)(s) :=
\begin{cases}
f(M_{n+1}(s)) & \textrm{if $M_n(s)$ is type-$i$}\\
0 & \textrm{o/w}
\end{cases}
\]

\begin{lemma}
\label{lem:inde}
For each $i$ and $n$, $f_{i, n+1}(M)$ is %not just $\mathcal{F}_{n+1}$-measurable but also 
$\mathcal{F}_{n}$-measurable.
\end{lemma}

\begin{proof} First fix notation
\begin{align*}
\sigma(M_{n+1}) &= \sigma\big(\set{M_{n+1}^{-1}(U_{\beta_k})
\mid \beta_k\in \mathsf{Sk}_k, U_{\beta_k} \in \mathcal{B}(\Real^k), k \geq 0}\big)\\
\sigma(M_{n}) &= \sigma\big(\set{M_{n}^{-1}(U_{\alpha_j})
\mid \alpha_j\in \mathsf{Sk}_j, U_{\alpha_j} \in \mathcal{B}(\Real^j), j \geq 0}\big);
\end{align*}
and for $\beta_k \in \mathsf{Sk}_i$, we write $\beta_k[\dagger]$ to mean the instantiation of $\beta_k$ by a $k$-vector of numerals $\dagger$; $\alpha_j[\dagger']$ has the same meaning.
Now for each $\beta_k[\dagger] = E[R']$ where $R'$ is the contractum of a type-$i$ redex $R$, there exist $\alpha_j \in \mathsf{Sk}_j$ and $j$-vector $\dagger'$ such that $\alpha_j[\dagger'] = E[R]$ of type $i$.
Moreover, for each $U_{\beta_k}$, there exists $U_{\alpha_j}$ such that $M_{n+1}^{-1}(U_{\beta_k}) = M_{n}^{-1}(U_{\alpha_j}) \in \sigma(M_n)$.
\end{proof}

Next we clearly have 
\(
f(M_{n+1}) = %\sum_{i=1}^3 f_{i, n+1}(M).
f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M),
\)
$\mu$-almost-surely.
(Unless otherwise stated, all equations and inequations between random variables are assumed to hold only $\mu$-a.s.)
Therefore 
\begin{align}
& \expect{f(M_{n+1}) \mid \mathcal{F}_n}
\nonumber \\ 
& = \expect{f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M)  \mid \mathcal{F}_n} 
\nonumber \\
& = \expect{f_{1, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{2, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{3, n+1}(M) \mid \mathcal{F}_n} 
\label{eqn:linear cond ex} \\
& = f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M)
\label{eqn:inde cond exp} \\
%& = f_{1, n+1}(M) + \int_I f() + f_{3, n+1}(M) \label{eqn:inde cond exp} \\
& \leq f(M_{n})[T_1] + f(M_{n})[T_2] + f(M_{n})[T_3]
\label{eqn:ts} \\
& = f(M_n). 
\label{eqn:partition} 
\end{align}
\Cref{eqn:linear cond ex} follows from the linearity of conditional expectation. 
\Cref{eqn:inde cond exp} is justified because $f_{i, n+1}(M)$ is $\mathcal{F}_n$-measurable (\Cref{lem:inde}).
Inequation~(\ref{eqn:ts}) follows from $f_{i, n+1}(M) \leq f(M_n)[T_i]$ for each $i$. 
Here we use the notation: given function $g : A \to \Real$ and $U \subseteq A$, $g[U] : A \to \Real$ denotes the function
\[
g[U](x) :=
\begin{cases}
g(x) & \textrm{$x \in U$}\\
0 & \textrm{o/w}
\end{cases}
\]
and write $T_i = \{s \in S \mid \textrm{$M_n(s)$ is type-$i$}\}$.
To see $f_{2, n+1}(M) \leq f(M_n)[T_2]$,
take $s \in T_2$. 
Then $f_{2, n+1}(M)(s) = f(E[\underline{a}])$ and $M_n = E[\tsample]$, for some $a \in [0, 1]$ and evaluation context $E$.
Hence 
\[
f_{2, n+1}(M)(s) \leq \int_I f(E[\underline{r}]) \, \mu_{leb}(\textrm{d} r)
\leq f(E[\tsample]) = f(M_n)[T_2](s).
\]
Finally, \Cref{eqn:partition} holds because $\{T_1, T_2, T_3, T_4\}$ is a partition of $S$.
}

@@

>>>>

\newcommand{\Real}{\mathbb{R}}
\newcommand\expect[1]{\mathbb{E}[#1]}
\newcommand\set[1]{\{#1\}}
\newcommand\dif{\mathrm{d}}
\newcommand\calF{\mathcal{F}}
\newcommand\calI{\mathcal{I}}
\newcommand\Leb{\mathrm{Leb}}
\newcommand\ndraw[2]{\#\mathrm{draw}_{#1}(#2)}

@@

\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}

@@

\begin{theorem} 
\label{thm:rankable gives supermartingale}
\changed[lo]{If a closed SPCF term is rankable by $f$, then $(f(M_n))_{n \in \omega}$ is a supermartingale adapted to the filtration $(\mathcal{F}_n)_{n \in \omega}$ where $\mathcal{F}_n = \sigma(M_1, \cdots, M_n)$.}
\end{theorem}

%\begin{proof}

\lo{We also need to show $f(M_n)$ is integrable.}

Given a closed term $M$ and a ranking function $f$ for it, define random variables on the probability space $S$ (where $s$ is a random variable) by
\begin{align*}
(M_n,s_n) & = \red^n(M,s) \\
y_0 & = 0 \\
y_{n+1} & = \min \{ k \mid k>y_n, M_k \text{ a value or of the form } E[\tY N] \}\\
M'_n & = M_{y_n} \\
X_n & = f(M'_n)
\end{align*}
and define a filtration $\mathcal{F}_n = \sigma(M_k, k \leq n)$ (i.e. all the samples used up to step $n$).

@@

@@

\input{ranking-supermartingales}
